---
title: "Formal Data Analysis: Factors Influencing IMDb Movie Quality"
author: "Group 10"
date: last-modified
number-sections: true
format: 
  html:
    embed-resources: true
editor_options:
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(broom)
library(kableExtra)
library(dplyr)
library(ggplot2)
library(scales)
library(sjPlot)

library(brglm2)
options(scipen = 999)
```

# Formal Data Analysis {#sec-fda}

Based on the exploratory analysis, we now fit a formal Generalised Linear Model (GLM) to investigate which movie characteristics are associated with a higher probability of receiving an IMDb rating above 7.

## Model Setup

```{r}
#| warning: false
#| message: false

# Prepare model data for logistic regression
setwd('C:/Users/16405/Documents')
df_clean <- read.csv('dataset10.csv')
model_data <- df_clean %>%
  mutate(
    # change the small sample into "Other"
    genre = forcats::fct_lump(factor(genre), prop = 0.05),
    # log votes, prevent the high number which caused coverage question.
    log_votes = log10(votes + 1), 
    is_high_rating = as.numeric(is_high_rating)
  ) %>%
  filter(!is.na(genre)) # Check Na

model_data$genre <- relevel(model_data$genre, ref = "Comedy")

# Check structure of model data
#str(model_data)
#summary(model_data)
```

The response variable for the formal analysis is `is_high_rating`, which equals 1 if a movie has a rating above 7 and 0 otherwise. Since the response is binary, a logistic regression model is used.

Based on the exploratory analysis, the predictors included in the model are `year`, `length`, `budget`, `log_votes`, and `genre`. A reference category is specified for `genre` so that the coefficients of the other genres can be interpreted relative to this baseline.

To improve model stability, small genre categories are combined into Other, and vote counts are log-transformed to reduce skewness. \## Logistic Regression Model

```{r}
#| warning: false
#| message: false

# Fit a full logistic regression model
glm_full <- glm(
  is_high_rating ~ year + length + budget + log_votes + genre,
  data = model_data,
  family = binomial(link = "logit"),
  method = "brglmFit" #Add some penalty to restrain extreme number.
)
# Display model summary
summary(glm_full)
```

The logistic regression results show that several predictors are significantly associated with the probability of a movie receiving a rating above 7.

In particular, higher budgets and greater audience engagement (`log_votes`) are positively associated with high ratings, while movie length shows a significant negative association. Genre is also an important predictor, with substantial differences in predicted odds across categories relative to the reference genre.

## Model Coefficients Table

```{r}
#| warning: false
#| message: false

tidy(glm_full, conf.int = TRUE, exponentiate = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kbl(
    col.names = c("Term", "OR", "SE", "z", "p", "CI low", "CI high"),
    caption = "Logistic regression odds ratios with 95% confidence intervals"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

table(model_data$genre, model_data$is_high_rating)
```

```{r}
## Predicted Probability Plot
#| warning: false
#| message: false

# Create a prediction grid
nd <- expand.grid(
  log_votes = seq(
    min(model_data$log_votes, na.rm = TRUE),
    max(model_data$log_votes, na.rm = TRUE),
    length.out = 100
  ),
  genre = levels(model_data$genre),
  year = median(model_data$year, na.rm = TRUE),
  length = median(model_data$length, na.rm = TRUE),
  budget = median(model_data$budget, na.rm = TRUE)
)

# Predict probabilities from the fitted model
nd$pred <- predict(glm_full, newdata = nd, type = "response")

# Plot predicted probabilities
ggplot(nd, aes(x = log_votes, y = pred, colour = genre, fill = genre)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Log(Votes)",
    y = "Predicted probability of rating > 7",
    colour = "Genre",
    fill = "Genre",
    title = "Predicted probability of high movie rating by audience engagement",
    subtitle = "Year, length and budget held fixed at their median values"
  ) +
theme_minimal()
```

To visualise the fitted logistic regression model, we plot the predicted probability of a movie receiving a rating above 7 across different levels of audience engagement (log_votes), while holding year, length and budget fixed at their median values. Separate curves are shown for each genre.

This plot provides a direct illustration of the model predictions. In general, the probability of receiving a high rating increases with audience engagement, although the level of this probability differs substantially across genres.

```{r}
# Coefficient plot with CIs
plot_model(
  glm_full,
  type = "est",
  transform = "exp",
  show.values = TRUE,
  value.offset = 0.25
)
```

The coefficient plot shows the estimated odds ratios and confidence intervals for each predictor. This gives a concise visual summary of the direction and strength of the associations identified in the logistic regression model.
