---
title: "Exploratory Data Analysis: Factors Influencing IMDb Movie Quality"
author: "Group 10"
date: last-modified
number-sections: true
format: 
  html:
    embed-resources: true
editor_options:
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false

# the lines above ensure that the R code in this block (along with any warning messages) will not be shown
# Load necessary libraries
library(ggplot2)
library(tidyr)
library(corrplot)
library(gridExtra)
library(readr)
library(knitr)
library(broom)
library(kableExtra)
library(scales)
library(sjPlot)
library(dplyr)
#Address conflicted
library(conflicted)
#Address the quasi-complete separation issue 
library(brglm2)
options(scipen = 999)
conflict_prefer("select", "dplyr")
# Define a harmonized color palette for consistency
# Orange represents 'Poor' (lower quality), Blue represents 'Good' (higher quality)
my_colors <- c("Poor (<=7)" = "#E69F00", "Good (>7)" = "#1b9e77")
```

```{r}
#| echo: false
#| warning: false
#| message: false
# Load and Clean Data
# read data(Users can choose a way to load a file)
#setwd("~/Documents/GitHub/DAS-Group-10/") # better use by different person 
#df <- read_csv("dataset10.csv")
##important: you need change it to your own path where your csv in 
##right-click your folder, ctrl+shift+C or "copy file path"
##remember to change "\" to "/"
df <- read_csv("https://raw.githubusercontent.com/Lee071502/DAS-Group-10/refs/heads/main/dataset10.csv")

```

```{r}
#| echo: false
#| warning: false
#| message: false
# Remove 'film_id' as it is a unique identifier with no predictive power
df <- df %>% select(-film_id)

# Check for missing values (NAs)
#print("Count of missing values per column:")
#print(colSums(is.na(df)))

# Drop rows with any missing values
df_clean <- df %>% drop_na()

# Create binary target: 1 if rating > 7 (High), 0 otherwise (Low)
df_clean <- df_clean %>%
  mutate(
    is_high_rating = ifelse(rating > 7, 1, 0),
    is_high_rating_factor = factor(is_high_rating, levels = c(0, 1), labels = c("Poor (<=7)", "Good (>7)")),
    genre = as.factor(genre) # Ensure genre is a categorical factor
  )

```

# Introduction {#sec-Intro}

The following analysis explores the internal characteristics of the movie dataset to identify potential predictors for our Generalized Linear Model (GLM). We specifically focus on whether a movie achieves an IMDb rating greater than 7.

# Exploratory Analysis {#sec-EA}

## Data Summary & Descriptive Statistics

To understand the central tendency and dispersion of our movie features, we generated a summary table for numerical variables. As presented in @tbl-summary, the dataset covers a wide range of years and budgets.

```{r}
#| echo: false
#| label: tbl-summary
#| tbl-cap: Descriptive Statistics for Numerical Movie Features 

# Create a summary table
summary_stats <- df_clean %>%
  select(year, length, budget, votes, rating) %>%
  summarise(across(everything(), list(
    Mean = ~mean(.x, na.rm = TRUE),
    SD = ~sd(.x, na.rm = TRUE),
    Min = ~min(.x, na.rm = TRUE),
    Max = ~max(.x, na.rm = TRUE)
  ))) %>%
  pivot_longer(everything(), names_to = c("Variable", "Statistic"), names_sep = "_") %>%
  pivot_wider(names_from = Statistic, values_from = value)

# Render table using kable
kable(summary_stats, digits = 2)
```

The diversity of movie genres is another key aspect of our dataset. @tbl-genre shows the count and percentage of movies within each genre category after cleaning the missing values.

```{r}
#| echo: false
#| label: tbl-genre
#| tbl-cap: Frequency and Percentage Distribution of Movie Genres

# Create a frequency table for Genre
genre_table <- df_clean %>%
  group_by(genre) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = paste0(round(Count / sum(Count) * 100, 1), "%")) %>%
  arrange(desc(Count))

# Render table
kable(genre_table)
```

By reviewing @tbl-summary, the variance of `votes` is exceptionally high compared to other predictors. This extreme dispersion indicates a "long-tail" distribution where a few extremely popular films receive a disproportionate amount of audience engagement, illustrating the need for a further transformation to stabilize the variable for the GLM. Furthermore, @tbl-genre highlights that Drama and Action are the  two most prevalent categories in our sample. In contrast, Romance represents the smallest segment of the sample. This distribution reflects the producer's focus on mainstream genres but suggests that the model's predictive power for specific genres like Romance may be more limited.

## Class Distribution

We examine the balance of our target variable (movie quality). As shown in @fig-dist, we can observe the frequency of movies categorized as "Good" versus "Poor".

```{r}
#| echo: false
#| label: fig-dist
#| fig-cap: Distribution of Movie Quality Classes based on IMDb Ratings
ggplot(df_clean, aes(x = is_high_rating_factor, fill = is_high_rating_factor)) +
  geom_bar() +
  scale_fill_manual(values = my_colors) +
  labs(x = "Rating Category", y = "Frequency", fill = "Category") +
  theme_minimal()
```

Based on @fig-dist, a significant class imbalance is observed in the target variable, with 'Poor' films ($Rating < 7$) outnumbering 'Good' films ($Rating > 7$) by nearly a factor of two. Because the model sees more examples of 'Poor' films, it might become better at identifying failures than predicting successes.

## Numerical Variables vs Target

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-box
#| fig-cap: Boxplots of Movie Length, Budget, and Votes by Quality Category
#| fig-align: center
#| fig-width: 10
#| fig-height: 8

# Analyzing Length, Budget, Votes, and Year using Boxplots
# A. Length: Highlighting the standardized 'Poor' movies vs. flexible 'Good' movies
p_length <- ggplot(df_clean, aes(x = is_high_rating_factor, y = length, fill = is_high_rating_factor)) +
  geom_boxplot(alpha = 0.8, outlier.color = "red") + 
  scale_fill_manual(values = my_colors) +
  labs(title = "Movie Length vs Quality", x = "", y = "Duration (Min)") + 
  theme_minimal() + theme(legend.position = "none")

# B. Votes: Addressing extreme right-skewness in both categories
p_votes <- ggplot(df_clean, aes(x = is_high_rating_factor, y = votes, fill = is_high_rating_factor)) +
  geom_boxplot(alpha = 0.8, outlier.color = "red") + 
  scale_fill_manual(values = my_colors) +
  labs(title = "Votes vs Quality", x = "", y = "Votes") + 
  theme_minimal() + theme(legend.position = "none")

# C. Year & D. Budget: Demonstrating weak correlation with quality
p_year <- ggplot(df_clean, aes(x = is_high_rating_factor, y = year, fill = is_high_rating_factor)) +
  geom_boxplot(alpha = 0.8) + scale_fill_manual(values = my_colors) +
  labs(title = "Year vs Quality", x = "") + theme_minimal() + theme(legend.position = "none")

p_budget <- ggplot(df_clean, aes(x = is_high_rating_factor, y = budget, fill = is_high_rating_factor)) +
  geom_boxplot(alpha = 0.8) + scale_fill_manual(values = my_colors) +
  labs(title = "Budget vs Quality", x = "") + theme_minimal() + theme(legend.position = "none")

# Combine plots into a 2x2 grid
grid.arrange(p_year, p_budget, p_length, p_votes, ncol = 2)

```

Analysis of the numerical predictors in @fig-box reveals that while `Year` and `Budget` show nearly identical distributions across quality classes, `Length` and `Votes` offer more significant insights. Notably, "Poor" movies exhibit a "Mediocrity Trap," with their durations tightly clustered around a standardized 100-minute mark, yet plagued by numerous extreme outliers (represented by the red markers in @fig-box). In contrast, "Good" movies—while also averaging near 100 minutes—display a much more flexible and varied distribution. Furthermore, because both the "Good" and "Poor" categories exhibit extreme right-skewness in their raw audience engagement metrics, we utilized a logarithmic transformation for the `Votes` variable. As shown in the red-highlighted outliers, this step was essential to stabilize the variance.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-votes-transformation
#| fig-cap: "Comparison of Votes Distribution: Raw vs. Log-transformed"
#| fig-height: 5
#| fig-width: 12

# Prepare Log-transformed Data
df_clean <- df_clean %>%
  mutate(log_votes = log1p(votes))

# A. Raw Votes: Showing extreme right-skewness with a long tail
p_raw <- ggplot(df_clean, aes(x = votes, fill = is_high_rating_factor)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = my_colors) +
  labs(title = "Original Votes Distribution", 
       subtitle = "Extreme right-skewness observed",
       x = "Number of Votes", y = "Density") +
  theme_minimal() + theme(legend.position = "top")

# B. Log-transformed Votes: Showing a more normalized, stable distribution
p_log <- ggplot(df_clean, aes(x = log_votes, fill = is_high_rating_factor)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = my_colors) +
  labs(title = "Log-transformed Votes", 
       subtitle = "Variance stabilized for GLM modeling",
       x = "Log(Votes + 1)", y = "Density") +
  theme_minimal() + theme(legend.position = "top")

# Combine side-by-side
grid.arrange(p_raw, p_log, ncol = 2)
```

To address the extreme right-skewness inherent in audience engagement metrics, we compared the raw distribution of votes against a log-transformed version. As shown in @fig-votes-transformation, the original `votes` variable is heavily compressed near zero with an expansive long tail, which would disproportionately bias our GLM. By applying a logarithmic transformation, we successfully achieved a more bell-shaped, stable distribution (Right), ensuring that the variance is consistent across both quality categories and making the data suitable for parametric modeling.

## Categorical Analysis of Genre

```{r}
#| echo: false
#| label: fig-genre
#| fig-cap: Proportion of Good vs. Poor Movies across Different Genres

# Proportional stacked bar chart for Movie Genre
ggplot(df_clean, aes(x = genre, fill = is_high_rating_factor)) +
  geom_bar(position = "fill") +
  coord_flip() +
  scale_fill_manual(values = my_colors)+
  labs(title = "Impact of Genre on Movie Quality",
       x = "Genre", y = "Proportion", fill = "Quality") +
  theme_minimal()
```

As illustrated in @fig-genre, genres such as Documentary and Short appear to have a significantly higher probability of being rated above 7.0 compared to Action or Drama films.

## Multi-collinearity Check & Correlation

```{r}
#| echo: false
#| label: fig-corr
#| fig-cap: Full 4x4 Correlation Matrix for Numerical Predictors  

# Extract numerical columns for correlation analysis
numeric_vars <- df_clean %>% 
  select(year, length, budget, votes, log_votes) 
#calculate correlation
cor_mat <- cor(numeric_vars, use = "complete.obs")
corrplot(cor_mat, 
         method = "color", 
         type = "upper", 
         addCoef.col = "black", 
         tl.col = "black", 
         tl.srt = 45,
         col = colorRampPalette(c("#E69F00", "white", "#1b9e77"))(200),
         title = "Feature Correlation Matrix",
         mar = c(0,0,1,0))# Adjust margin to fit title
# Final Step: Export cleaned data for the GLM modeler
#write.csv(df_clean, "cleaned_movie_data.csv", row.names = FALSE)


```

As shown in @fig-corr, no predictors exhibit a correlation higher than 0.7, confirming that we can include all variables in the GLM.

# Summary of Analysis {#sec-sum}

Our exploratory analysis shows several patterns in the data. First, there is a clear class imbalance: “Poor” films appear about twice as often as “Good” films, which creates a strong baseline for prediction. We also find that budget and production year have little relationship with film ratings. However, film length shows an interesting pattern. Lower-rated films tend to cluster around 100 minutes, while higher-rated films have a wider range of lengths.

In addition, the number of audience votes is highly right-skewed, so we applied a log transformation to stabilize the variance. After this transformation, higher-rated films still show consistently higher audience engagement. By combining these findings with differences across genres, we gain a clearer understanding of the factors related to movie quality and build a stronger foundation for our GLM model.

# Formal Data Analysis {#sec-fda}

Based on the exploratory analysis, we now fit a formal Generalised Linear Model (GLM) to investigate which movie characteristics are associated with a higher probability of receiving an IMDb rating above 7.

## Model Setup
```{r}
#| echo: false
#| warning: false
#| message: false

# Prepare model data for logistic regression
model_data <- df_clean %>%
  mutate(
    # change the small sample into "Other"
    genre = forcats::fct_lump(factor(genre), prop = 0.05),
    # log votes, prevent the high number which caused coverage question.
    log_votes = log(votes + 1), 
    is_high_rating = as.numeric(is_high_rating)
)

model_data$genre <- relevel(model_data$genre, ref = "Comedy")
```

The response variable for the formal analysis is `is_high_rating`, which equals 1 if a movie has a rating above 7 and 0 otherwise. Since the response is binary, a logistic regression model is appropriate.

Based on the exploratory analysis, the predictors included in the model are `year`, `length`, `budget`, `log_votes`, and `genre`. The reference category for `genre` is set to Comedy, so that the coefficients of other genres can be interpreted relative to this baseline.

To improve model stability, rare genre categories are grouped into an "Other" category, and vote counts are log-transformed to reduce skewness. The final dataset contains `r nrow(model_data)` observations after preprocessing.

## Logistic Regression Model
To investigate the determinants of a movie's likelihood of receiving a high rating, we employed a Logistic Regression model. Given the potential for quasi-complete separation in categorical data, we utilized the Firth’s penalized maximum likelihood estimation (via the brglmFit method) to ensure parameter stability and reduce bias.
$$
\ln\left( \frac{P}{1 - P} \right) = \beta_0 + \beta_1 \text{year} + \beta_2 \text{length} + \beta_3 \text{budget} + \beta_4 \text{log\_votes} + \sum_{i=1}^{n} \gamma_i \text{genre}_i
$$

Where:

* **$P$**: The probability that the outcome variable `is_high_rating` equals 1 (representing a high-rated movie).
* **$\text{log\_votes}$**: The natural logarithm of the vote count, used to normalize the distribution and reduce the impact of extreme outliers.
* **$\text{genre}$**: A categorical predictor where rare genres have been grouped into an "Other" category to improve model stability.
* **$\beta_i$**: The estimated coefficients for the continuous and discrete numerical predictors.
* **$\gamma_i$**: The estimated coefficients for the specific movie genre categories.
* **$\ln\left( \frac{P}{1 - P} \right)$**: The log-odds (logit), which represents the link between the linear predictors and the probability of the outcome.
```{r}
#| echo: false
#| warning: false
#| message: false

# Fit a full logistic regression model
glm_test <- glm(
  is_high_rating ~ year + length + budget + log_votes + genre,
  data = model_data,
  family = binomial(link = "logit"),
  method = "brglmFit" #Add some penalty to restrain extreme number.
)
# Display model summary
summary(glm_test)
```

An initial model including `year` was fitted. However, as `year` was not statistically significant (p > 0.05), it was removed to improve model parsimony.

So our final model would be:
$$
\ln\left( \frac{P}{1 - P} \right) = \beta_0 + \beta_1 \text{length} + \beta_2 \text{budget} + \beta_3 \text{log\_votes} + \sum_{i=1}^{n} \gamma_i \text{genre}_i
$$
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-Glmsummary

# Fit a full logistic regression model
glm_full <- glm(
  is_high_rating ~ length + budget + log_votes + genre,
  data = model_data,
  family = binomial(link = "logit"),
  method = "brglmFit" #Add some penalty to restrain extreme number.
)
# Display model summary
summary(glm_full)
```

@tbl-Glmsummary indicates that several predictors are significantly associated with the probability of receiving a rating above 7.

Higher budgets and greater audience engagement (`log_votes`) increase the likelihood of a high rating, while longer movie duration reduces it. 

Genre also plays a substantial role: compared with Comedy, Documentary films have markedly higher odds of receiving high ratings, whereas Animation and Drama are associated with lower odds.

## Model Coefficients Table
```{r}
#| label: tbl-or-table
#| warning: false
#| message: false
#| echo: false

tidy(glm_full, conf.int = TRUE, exponentiate = TRUE) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kbl(
    col.names = c("Term", "OR", "SE", "z", "p", "CI low", "CI high"),
    caption = "Logistic regression odds ratios with 95% confidence intervals"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
#just check the imbalance of genre
#table(model_data$genre, model_data$is_high_rating)
```
@tbl-or-table reports the estimated odds ratios, standard errors and confidence intervals for all predictors in the final logistic regression model.

## Predicted Probability Plot
```{r}
#| label: fig-pred-plot
#| fig-cap: prediction probability plot
#| warning: false
#| message: false
#| echo: false

# Create a prediction grid
nd <- expand.grid(
  log_votes = seq(
    min(model_data$log_votes, na.rm = TRUE),
    max(model_data$log_votes, na.rm = TRUE),
    length.out = 100
  ),
  genre = levels(model_data$genre),
  length = median(model_data$length, na.rm = TRUE),
  budget = median(model_data$budget, na.rm = TRUE)
)

# Predict probabilities from the fitted model
nd$pred <- predict(glm_full, newdata = nd, type = "response")

# Plot predicted probabilities
ggplot(nd, aes(x = log_votes, y = pred, colour = genre, fill = genre)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Log(Votes)",
    y = "Predicted probability of rating > 7",
    colour = "Genre",
    fill = "Genre",
    title = "Predicted probability of high movie rating by audience engagement",
    subtitle = "Length and budget held fixed at their median values"
  ) +
theme_minimal()
```

To visualise the fitted model, @fig-pred-plot shows the predicted probability of a movie receiving a rating above 7 across different levels of audience engagement (`log_votes`), while holding other variables fixed at their median values.

We also observe that the probability of a high rating increases with audience engagement. The approximately parallel curves indicate that the effect of `log_votes` is consistent across genres, with differences primarily reflected in baseline probabilities.

```{r}
#| label: fig-coef-plot
#| echo: false
#| warning: false
#| fig-cap: Coefficient plot with CIs
plot_model(
  glm_full,
  type = "est",
  transform = "exp",
  show.values = TRUE,
  value.offset = 0.25
)
```

@fig-coef-plot presents the estimated odds ratios and confidence intervals for each predictor.

The plot highlights the magnitude of effects on the odds scale. Documentary films have odds of high ratings more than 10 times that of Comedy, whereas Animation and Drama have substantially lower odds. Continuous predictors such as budget and `log_votes` increase the odds of high ratings, while movie length reduces them.

## Conclusion

This study identifies audience engagement, budget, and genre as key determinants of high movie ratings. Movies with higher vote counts and larger budgets are significantly more likely to receive ratings above 7, whereas longer duration is associated with reduced likelihood.

Substantial differences across genres persist after controlling for other variables, highlighting the independent role of genre in shaping perceived movie quality. 

Overall, the fitted logistic regression model offers a stable and interpretable framework for explaining variation in IMDb ratings.
